\section{Plan}

Our project proceeds through several structured phases, including implementation, experimentation, analysis, and documentation. The primary goal of the remaining project timeline is to complete the experimental pipeline for extracting safety and utility representations LLMs, evaluate their geometric relationship, and finalize both the empirical results and written report. The project will be developed collaboratively by the six group members, with responsibilities distributed across both implementation and writing tasks.

During the first phase of the project, we focus on finalizing the experimental infrastructure and ensuring that the datasets, models, and extraction pipelines are working correctly. We will complete the activation extraction pipeline using the Gemma and Llama model checkpoints, collecting hidden states from the final layers of the network for prompts drawn from the AdvBench and HelpSteer datasets. These datasets allow us to define the behavioral poles of safety (refusal of harmful prompts) and utility (following benign, helpful instructions). In addition, we will verify that the data preprocessing and activation logging steps produce stable and reproducible outputs.

Following the data preparation stage, we will implement the two primary direction extraction techniques described in our proposal: DoM actSVD. The DoM method will be used as a baseline approach to extract a single refusal direction representing safety behavior by comparing the average activations between harmful and harmless prompts. In parallel, we will implement actSVD in order to identify higher-dimensional safety and utility subspaces within the activation space. Once both extraction methods are implemented, we will compare their outputs to determine whether they isolate similar behavioral directions.

After the extraction phase, we will conduct the main set of experiments evaluating the relationship between safety and utility representations. This includes computing cosine similarity between the extracted vectors, measuring subspace overlap, and performing cross-intervention experiments. In these experiments we will ablate the safety direction and measure the effect on utility benchmarks, and conversely remove the utility direction and measure the effect on safety behaviors. These intervention experiments allow us to test whether safety and utility behave as causally separable concepts within the activation space.

By the Project Progress Discussion on April 10, we plan to have the majority of the experimental implementation completed. Specifically, the activation extraction pipeline, the DoM baseline, and the actSVD implementation will all be functional. We also aim to have preliminary experimental results, including early measurements of cosine similarity and initial intervention tests. This milestone will allow us to identify potential issues early and refine our experimental setup before finalizing the project.

Following the progress discussion, the remaining time will be dedicated to completing the final experiments, analyzing the results, and preparing the final report and presentation. During this stage we will run additional experiments across multiple model checkpoints to verify the consistency of our findings, generate visualizations such as PCA projections of activation spaces, and summarize the empirical patterns observed in the data. The final stage will also include writing the discussion and conclusion sections, polishing the figures, and preparing the final presentation slides.

The workload for the project will be distributed evenly among the six group members. For the final report, each member will contribute to both the experimental implementation and the written documentation. Individual responsibilities will be assigned by dividing the major components of the project, including activation extraction, direction extraction methods, experimental evaluation, visualization, and report writing. This approach ensures that each group member contributes both technically and academically to the project. Additionally, all members will participate in reviewing the results, debugging experiments, and refining the final report to maintain consistency and clarity.

Overall, this plan ensures steady progress toward the final deliverables while allowing sufficient time for experimentation, analysis, and collaborative writing. By completing the majority of the implementation before the progress checkpoint and reserving the final period for evaluation and refinement, we aim to produce a well-supported empirical study of the relationship between safety and utility representations in LLM activation spaces.

Finally, this plan represents our current proposal, and some implementation details or experimental procedures may change as the project progresses, as an example, we are trying to figure out if we may use the causal inner product as presented in \citet{park2023LRH} to gain greater insight into causal independence of safety and utility, and test larger models should we find that safety and utility are not causally separable for more substantive findings, as larger LLMs tend to be more separable.
