\section{Appendix}
\subsection{Sample Size Lower Bound}
A possible improvement to extract the utility direction is to see the difference of usefulness=4 and the usefulness mean. 
This (Obviously) requires the mean of the activations on varied helpfulness scores on HelpSteer \citep{wang-etal-2024-helpsteer}. 
To do this we first take a sample of HelpSteer to estimate the mean, in this section we aim to find a lower bound for the sample size 
to get a good estimate of the mean.

\begin{lemma}
Let $G = \{g_1, \dots, g_5\}$ be the set of groups, and choose a group $g_k \in G$ 
that has the distribution $P_{g_k}$. Then let $x_1, \dots, x_n \stackrel{\mathrm{iid}}{\sim} P_{g_k}$ 
be samples from $g_k$, and let $a_i = f(x_i)$ be the activations produced by passing 
each $x_i$ through the network. Then $a_1, \dots, a_n$ are i.i.d
\end{lemma}

\begin{proof}
The $a_i$ are identically distributed because each $x_i$ is drawn from the same 
distribution $P_{g_k}$ and each $a_i$ is produced by passing forward i.i.d samples on the
exact same model, therefore each $a_i$ is drawn from the same distribution of activations. 
Further, computing $f(x_i)$ shares no state whatsoever with computing $f(x_j)$ for
$i \neq j$ and thus $a_i$ and $a_j$ are independent \qedhere
\end{proof}

\textbf{Note}: For the previous lemma to hold we require that each activation is sampled the same way, we extract activations from 
the last token so this was omitted as it is not necessary to list that condition for our case.

\begin{theorem}
Let $\epsilon, \delta > 0$ then we sample $n_k=p_k N$ i.i.d samples from each group $g_k \in G$, and $p_k$
is the proportion of $g_k$ in HelpSteer. We pass the samples and construct an estimator of the mean
$\hat{ \mu}=\sum_{k=1}^{5}p_k \bar{a} _k$. Then if $N \geq \frac{tr(\Sigma)}{\delta \epsilon ^2}$ we have that
$P(||\mu - \hat{\mu}|| < \epsilon) \geq 1 - \delta$
where $\Sigma = \sum_k p_k \Sigma_k$
\end{theorem}
\begin{proof}
By lemma we have that the activations from each group $g _k \in G$ are i.i.d, so 
we have that $\mathbb{E} [\bar{a} _k] = \mu _k$ and then using the linearity of expectation we get that


$\mathbf{E}[\sum _k p_k \bar{a} _k]=\sum_k p_k \mathbf{E}[\bar{a} _k] = \sum_k p_k \mu _k=\mu$ so the
mean we have is not biased. Next we want to find the variance of $\hat{\mu}$, we find this by using
our lemma, we have that $\mathbf{Var}(\bar{a} _k)=\frac{\Sigma _k}{n _k}$, further, 
$\mathbf{Var}(\hat{\mu})= \sum _k \mathbf{Var}(p_k \bar{a _k})= \sum _k p_k^2\frac{\Sigma _k}{n _k}$ using
the definition of $\hat{\mu}, \bar{a}$ and basic property of the variance. Then we substitute $n_k = p_k N$
and get 

$\sum _k p_k \frac{\Sigma _k}{N}=\frac{\bar{\Sigma}}{N}$ by moving the costant out and taking the sum.
So, $\mathbf{Var}(\hat{\mu})=\frac{\bar{\Sigma}}{N}$, finally we apply Chebyshev's inequality
and put $\mathbf{Var}(\hat{\mu})=\frac{\bar{\Sigma}}{N}$ and get that 


$p(|| \hat{\mu} - \mu || < \epsilon ) \geq 1 - \frac{\bar{\Sigma}}{N \epsilon ^2}$,
and then solving for $N$ gives us $N \geq \frac{tr(\bar{ \Sigma })}{\delta \epsilon ^2}$


\end{proof}
