\section{Contribution Statement}
This project was a collaborative effort by the six group members. All group members 
contributed towards writing both the proposal and the final reports, the making of the
slides for our presentations and just overall making this project happen. 

The early data analysis was done by Zakeriya and David, this helped us gain better insight about the data
that we are dealing with as well as forming our initial hypothesis about what we should expect from each
experiment, and most importantly it aided designing and implementing our experiments by telling us what
to keep in mind in our analysis as well as what to account for during our experiments.

The early literature review was done by Will, which established the theoretical foundations of our 
experiments, namely, ActSVD and DoM. Which allowed us to then modify these existing methods to better 
fit our experiments and what we intend to do.

The brainstorming and discussing how the experiments should specifically be set up was done
by Khalid and Vivek. This made sure that the experiments were reasonable and that our logic was sound.
This was crucial for obtaining results that do tell something meaningful and designing the experiments.

The experiment pipeline was designed and implemented by Khalid, Vivek and Mohammed. Khalid implemented
and ran the DoM part as well as the extracting the activations part, 
Vivek implemented and ran the ablations part and Mohammed implemented and ran the SVD parts, in particular,
ActSVD and contrastive SVD. This was done in order to ensure that our experiments can be easily replicated and to ensure that our code and 
repository in general are neat, easy to read and navigate through.

Finally, the results analysis was done by Khalid, Vivek and Mohammed. Where again, Khalid focused on the 
DoM part, Vivek focused on the ablations part and Mohammed focused on the SVD part. This helped us form what
we found interesting from our results. 

Despite the seemingly clear division of work, throughout the project each part of our work was brought up
in a groupchat with all members to discuss.

All of our code and a link to a Google Drive containing our extracted activations 
can be found on my Github repository \href{https://github.com/khalidraisi/safety-utility-disentanglement.git}{here}.