\section{Discussion and Conclusion}
Our results indicate that the entanglement of safety and utility is layer dependent,
where in the earlier layers of the model they are aligned/anti-aligned. However, at the
mid-late layers we often find that they become near orthogonal. And generally speaking,
this is consistent across the different subspace extraction methods we use, namely,
DoM, ActSVD and contrastive SVD. Furthermore, we found that ablations also do preserve
similar performance on the non-ablated property. Both of which suggest that our 
interpretation of the Activation space could be indeed true and this is 
not a mere coincidence, another possibile way to test this is by establishing the 
"reverse" of this relation, i.e could a sparse autoencoder from a dataset containing
jumbled up refusal and useful inducing behaviour reproduce the correct sets? If so 
this could be more evidence that they are separable. This nevertheless motivates further
work to be done in this to potentially establish causal independence as per
\citet{park2023LRH} framework establishing a more rigorous relationship between 
safety and utility, this work only produces empirical evidence for but does not 
demonstrate thoroughly, since we examine the behaviour and not the inherent
representation space. 

Our experiments focus on smaller models, so these findings may not be true for larger
models or models with different architectures. Also refusal filtering is another 
limitation, which do tend to let through refusal behaviour or they capture non-refusal
behaviour. Most importantly however, is that obtaining a usefulness subspace is inherently
trickier than the refusal subspace, as it is not as clearly defined. It's also harder to
work with since there is less of a contrast between it and the average LLM behaviour compared
to refusal behaviour. To address this we also propose another way of obtaining
a usefulness subspace, whereby one would center it at the mean of usefulness instead of
centering at the average LLM behaviour. This also nicely sets up work in examining the utility
subspace and answer if usefulness and uselessness are at all related or not. We l lay 
some of the prelimenary work for this in \ref{proof}.

We presented an analysis comparing the safety and utility subspaces extracted using 
multiple different methods. And then targetting our interventions at particular layers
examining causal independence of safety and utility in instruction tuned LLMs.