\section{Experimental Setup}
We run 500 prompts from each dataset, then run a forward pass, obtain the activations
and then perform our experiments and analysis (DoM, ActSVD, Contrastive SVD and LocateLayers).
Finally we ablate and examine the post ablation performance. In this section we discuss
our datasets and models.
\subsection{Models}
We experiment on five instruction-tuned models covering four distinct
architectural designs: Gemma 3 \citep{gemma_2025}, Gemma 4
\citep{gemma4}, Llama 3.2 \citep{grattafiori2024llama3herdmodels}, and
Qwen 3.5 \citep{qwen3.5}. 
Gemma 3 and Gemma 4 use hybrid local/global
attention transformers, and Gemma 4 introduces
Per-Layer Embeddings on top of that. Further, Llama 3.2 is a standard decoder-only transformer
with grouped-query attention, and Qwen 3.5 uses a hybrid architecture
interleaving linear attention with full attention.
Table~\ref{tab:models} summarizes the models we evaluate. information about the used
chat templates can be found in section \ref{app:Templates} of the appendix.

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrr}
\hline
Model & Params & $d$ & $L$ \\
\hline
Gemma-3-1B-IT          & 1B & 1152 & 26 \\
Gemma-3-4B-IT          & 4B & 2560 & 34 \\
Gemma-4-E2B-IT         & 2B & 1536 & 35 \\
Llama-3.2-3B-Instruct  & 3B & 3072 & 28 \\
Qwen-3.5-4B            & 4B & 2560 & 32 \\
\hline
\end{tabular}
\caption{Models evaluated. $d$ is the hidden dimension; $L$ is the number of transformer blocks.}
\label{tab:models}
\end{table}
\subsection{Datasets}
We would like to obtain directions for refusal and useful behaviour, to do this we use
AdvBench \citep{zou2023gcg} and Helpsteer \citep{wang-etal-2024-helpsteer}. 
For AdvBench we simply pick the first 500 prompts, to then use to induce refusal behaviour.
We use Helpsteer, but we filter to get 500 prompts with the maximum helpfulness score. 
Finally, we Alpaca \citep{alpaca} where we also pick the first 500 prompts, this is just 
done in order to have a certain baseline for the model's behaviour on average. Further
details about the prompts can be found in part \ref{app:datasets} of the appendix.
\subsection{Evaluations}
We evaluate the extracted directions and subspaces along three axes.
\textbf{Geometric metrics} (cosine similarity for DoM directions and
principal angles for ActSVD and Contrastive SVD subspaces) are
reported per layer to characterize the relationship between safety
and utility, with a permutation null (1000 iterations of shuffled
labels) accompanying the cosine similarities.
\textbf{Layer selection} picks the safety and utility layers by class-separability AUROC, projecting
activations onto the candidate direction and computing AUROC against
the relevant binary label. Behavioral metrics measure the effect of
ablation through refusal rate on AdvBench prompts and utility on
held-out HelpSteer prompts, with the off-diagonal cases (refusal
after utility ablation, utility after safety ablation) quantifying
causal separability. All figures are added to section \ref{sec:figures} of the appendix.