\section{Methods}

We work with three datasets corresponding to three behavioral
categories: Alpaca \cite{alpaca} (harmless instructions),
AdvBench \cite{zou2023gcg} (harmful prompts), and the
helpfulness-rated-4 subset of HelpSteer
\cite{wang-etal-2024-helpsteer} (utility prompts). For each
prompt $p \in P$ in any of these three sets, we apply the model's
chat template, run a forward pass, and save the activation at the
final token of every transformer block. This gives a set of
matrices $A_n \in \mathbb{R}^{|P| \times d}$ where $n$ indexes the
layer ($n=0$ is the embedding). The direction and subspace
extraction methods below all operate on these $A_n$.


\subsection{Difference of Means (DoM)}
Following Arditi et al.~\cite{arditi2024refusal}, we summarize each
behavior at layer $n$ as a single direction. Let
$\mu^{(n)}_{\text{harmless}}$, $\mu^{(n)}_{\text{harmful}}$, and
$\mu^{(n)}_{\text{utility}}$ denote the column means of $A_n$ over
harmless prompts (Alpaca), harmful prompts (AdvBench), and utility
prompts (the HelpSteer rated-helpfulness-4 subset) respectively.
We define the safety and utility directions as
\[
v^{(n)}_{\text{safety}} = \mu^{(n)}_{\text{harmful (AdvBench)}}
                          - \mu^{(n)}_{\text{harmless (Alpaca)}},
\]
\[
v^{(n)}_{\text{utility}} = \mu^{(n)}_{\text{utility (HelpSteer)}}
                          - \mu^{(n)}_{\text{harmless (Alpaca)}},
\]
and quantify entanglement at depth $n$ via the cosine similarity
between them. We retain the sign of the cosine: a value near zero
indicates orthogonality, while strongly positive or strongly
negative values indicate that safety and utility are encoded along
the same or opposing axes.

To distinguish the observed cosines from what would be expected
under no relationship between the two directions, we run a
permutation test at each layer. We pool all harmful (AdvBench) and
utility (HelpSteer) prompts and resample two random groups of
equal size from this pool 1000 times, computing the cosine between
random ``safety-like'' and ``utility-like'' directions for each
draw. The mean and central 95\% interval of this null distribution
are overlaid on every reported cosine curve.


\subsection{Activation Aware SVD (ActSVD)}
DoM compresses each behavior to a single direction. ActSVD,
following Wei et al.~\cite{wei2024brittleness}, generalizes this
to a rank-$r$ subspace. We mean-center each activation matrix
using the harmless baseline drawn from Alpaca,
\[
D^{(n)}_{\text{safety}} = A_n^{\text{AdvBench}}
   - \mathbf{1}\,\mu^{(n)}_{\text{Alpaca}}{}^{\!\top},
\]
\[
D^{(n)}_{\text{utility}} = A_n^{\text{HelpSteer}}
   - \mathbf{1}\,\mu^{(n)}_{\text{Alpaca}}{}^{\!\top},
\]
and take the top-$r$ left singular vectors of each as orthonormal
bases $U^{(n)}_{s},\, U^{(n)}_{u} \in \mathbb{R}^{d \times r}$. We
fix $r=10$ throughout. Subspace entanglement is summarized by the
mean squared cosine of principal angles,
$\frac{1}{r}\sum_{i} \cos^2 \theta_i$, computed from the singular
values of $U^{(n)\top}_{s} U^{(n)}_{u}$. The score lies in $[0,1]$,
where $0$ indicates orthogonal subspaces and $1$ indicates
identical subspaces. We additionally report the largest
principal-angle cosine $\cos \theta_1$ to expose any single shared
direction that the average might mask.

\subsection{Contrastive SVD}
To address confounders such as text length or syntax we also use contrastive SVD,
where we perform SVD on the difference vectors of contrastive pairs.
For safety, each AdvBench prompt $x_i^{\text{harm}}$ is paired with a content-matched
harmless rewrite $x_i^{\text{safe}}$ generated by \texttt{gpt-5-nano}, preserving
verb, structure, and length while swapping only the harmful target. At each
layer $\ell$ we form
$D_\ell = [\,h_\ell(x_i^{\text{harm}}) - h_\ell(x_i^{\text{safe}})\,]_{i=1}^{n}$
and take the top $r$ left singular vectors of $D_\ell^{\top}$ as the
contrastive safety subspace $U_\ell^{\text{con-safe}}$. For utility we apply the
same construction to HelpSteer prompts rated helpfulness$\,{=}\,4$ vs.\
helpfulness$\,{=}\,0$.


\subsection{Locating Safety and Utility Layers}
The geometric measures above produce one score per layer per
model. To identify which layers most cleanly separate the two
behaviors, we rank layers by two criteria. For separability
under DoM, we project held-out activations onto the per-layer
direction $v^{(n)}$ and compute the AUROC of the resulting scalar
for distinguishing the contrast pair (harmful vs.\ harmless for
safety, utility vs.\ harmless for utility). For ActSVD, we use
the inverse of the per-layer subspace overlap score, so that
lower overlap maps to higher separation. Layers ranked highly by
both criteria are simultaneously predictive of the target behavior
and geometrically separated from the complementary one.

\subsection{Subspace Ablations}
To test whether the geometric findings have causal force, we
intervene on the residual stream during generation and measure
behavioral change. We register a forward hook on a single
transformer block that, at every token position during both
prefill and decoding, projects out one or more unit directions $v$
from the hidden state via $h \leftarrow h - (h \cdot v)\, v$. The
hook acts only at the specified block; downstream blocks may
re-introduce the direction, but our analysis targets the layer
that maximizes separability rather than network-wide ablation.

The ablation layer is selected per direction by AUROC. For each
layer $n$ we project held-out activations onto the unit-normalized
DoM direction $v^{(n)}_{\text{safety}}$ and compute the AUROC of
the resulting scalar for distinguishing harmful from harmless
prompts; the layer maximizing this AUROC is the safety ablation
site $L_{\text{safety}}$. The utility ablation site
$L_{\text{utility}}$ is selected analogously using the
utility-versus-harmless contrast. Layer 0 (embedding) and the
final block are excluded from the search to avoid degenerate
picks.

We compare five conditions: (i) \emph{baseline} (no
intervention); (ii) \emph{ablate-safety}, projecting out
$v_{\text{safety}}$ at $L_{\text{safety}}$; (iii) \emph{ablate-
utility}, projecting out $v_{\text{utility}}$ at $L_{\text{utility}}$;
(iv) \emph{ablate-both}, applying both hooks simultaneously; and
(v) \emph{ablate-random}, a control that projects out a random
unit vector at $L_{\text{safety}}$. The random condition matches
the scope of the safety intervention, so any behavioral change
attributable to ``loss of one direction'' should appear here too;
differences between ablate-safety and ablate-random isolate the
content of the safety direction from the act of projection itself.

We evaluate each condition on two benchmarks. AdvBench measures
safety: we generate a short continuation for each harmful prompt
and compute the refusal rate. HelpSteer measures utility: we
generate longer responses for held-out prompts and score each on
a 1--5 helpfulness scale using GPT as a judge. Together these
benchmarks let us read causal independence directly --- ablating
safety should drop AdvBench refusal rate while leaving HelpSteer
helpfulness near baseline if the two behaviors are causally
separable, and the reverse for utility ablation.
