\section{Results}
\subsection{Subspace Geometry}
We characterize the internal geometry of safety and utility
representations using three methods: Difference-of-Means
(DoM), which extracts a single signed direction per behavior per layer;
ActSVD, which compares rank-$r=10$ subspaces of safety and utility
activations via principal angles; and ContrastiveSVD, which repeats
the subspace comparison on difference vectors between matched prompt
pairs, controlling for prompt-level confounders such as length and
syntactic form. All three methods are applied to all five models,
with the exception of ContrastiveSVD, which we run on Llama-3.2-3B
only~\ref{sec:figures-actsvdcsvd}. Across methods and models, a
consistent picture emerges: safety and utility occupy largely
orthogonal directions and subspaces in the residual stream, with
entanglement concentrated in early layers and near-perpendicular
separation emerging in the final third of each network.

The ActSVD overlap profiles for all five models share a common
depth structure. At layer 0 (the embedding), the overlap score is
exactly 1.0 for every model by construction, since both subspaces
are extracted from the same token distribution. Overlap then drops
sharply after the first transformer block and continues declining
monotonically through the middle layers, reaching its minimum in
the final third of each network. This pattern holds without
exception across all five architectures, suggesting that the
progressive separation of safety and utility representations is a
general property of instruction tuned transformers rather than an
artifact of any particular model family or scale.

Llama-3.2-3B exhibits the cleanest separation of all five models
under ActSVD. Its overlap score falls from 1.0 at layer 0 to
approximately 0.21 by layer 1, and continues declining steadily
through the middle layers, reaching values below 0.10 by layer 20
and achieving its minimum of approximately 0.04 at layer 28, the
penultimate layer of the network. The top principal-angle cosine
follows a similarly smooth monotone decline, falling from 1.0 at
the embedding to approximately 0.49 at layer 28. Critically, this
decline is smooth and uninterrupted --- there are no transient
recoveries toward high alignment at any depth --- and the minimum
top cosine of 0.49 means that even the single most-shared direction
between the safety and utility subspaces falls well short of
cosine 1 in late layers. The orthogonality is therefore distributed
across the full rank-10 subspace rather than being masked by a
single dominant shared axis.

The two ContrastiveSVD variants on Llama-3.2-3B replicate and
strengthen this finding. The utility-contrast variant shows an overlap score of
approximately 0.21 at layer 1, which then plateaus briefly before
declining to around 0.05 by layer 28. The paired-contrast safety
variant (paired contrastive safety subspace versus utility subspace)
follows a similar trajectory but starts somewhat higher at
approximately 0.30 at layer 1 and declines more sharply, reaching
values below 0.15 by layer 4 and stabilizing near 0.10 through
the remainder of the network. Both variants achieve their minimum
overlap in the final layers, consistent with ActSVD. Importantly,
both ContrastiveSVD overlap profiles are at most as large as the
corresponding ActSVD profile at every layer, and in the late layers
they are slightly lower. Since ContrastiveSVD controls for
prompt-level confounders that ActSVD cannot, this reduction confirms
that the separation observed under ActSVD is not inflated by shared
surface features such as text length or syntactic structure. The
safety and utility subspaces are genuinely disentangled in the
residual stream of Llama-3.2-3B, not merely artefacts of the
dataset construction.

The top principal-angle cosine under both ContrastiveSVD variants
reinforces this conclusion. For the utility-contrast variant it
declines from 1.0 at layer 0 to approximately 0.54 at layer 28,
with a local dip to around 0.62 near layer 15 before continuing
its decline. For the paired-contrast variant it falls from 1.0 to
approximately 0.56 at layer 28, with a local minimum near 0.58 at
layer 11. Neither variant recovers toward high alignment at any
depth. The smoothness of both profiles, and their close agreement
with each other, indicates that the geometric finding is robust
to the specific choice of contrastive pairing strategy.

\subsection{DoM}
\label{sec:results-dom}

Across post-embedding layers, the mean signed cosine between the
DoM safety and utility directions stays close to zero: $+0.10$ for
Gemma-3-1B, $-0.24$ for Gemma-3-4B, $+0.17$ for Gemma-4-E2B, $+0.06$
for Llama-3.2-3B, and $+0.10$ for Qwen-3.5-4B. The sign indicates
whether the directions tend to point with each other ($+$) or
against each other ($-$); the magnitude indicates how strongly. None
of the five models exhibits a sustained alignment in either
direction.

The picture remains the same when looking at individual layers.
Several achieve near-perfect orthogonality: layer 8 of Gemma-3-4B
($\cos = +0.04$), layers 11 and 15 of Gemma-4-E2B ($\cos = +0.01,
+0.02$), layers 17--20 of Llama-3.2-3B ($-0.02 \le \cos \le +0.02$),
and layers 6--9 of Qwen-3.5-4B ($-0.01 \le \cos \le +0.01$). When
measured against a random-pair null distribution, no layer in any
model exceeds the upper null bound, while a substantial fraction
fall below the lower bound: 17 of 29 layers in Llama-3.2-3B and 15
of 33 in Qwen-3.5-4B. For these two models, DoM safety and utility
are \emph{more} orthogonal than chance at most depths.

Gemma-3-4B is the partial exception. Its safety and utility
directions are strongly \emph{anti-aligned} at layer 11
($\cos = -0.90$), and weakly aligned at layer 6 ($\cos = +0.62$).
The negative cosine at layer 11 means refusal and helpfulness are
encoded along opposite axes there: one direction simultaneously
pushes toward refusal and away from helpful response. This
phenomenon is layer-localized rather than global --- 32 of the
model's 35 layers still fall within the null band --- but it shows
that not all layers separate the two behaviors orthogonally even
when the model average suggests they do.

Ranking the five models from most to least orthogonal under DoM
(by mean $|\cos|$): Qwen-3.5-4B (0.10), Llama-3.2-3B (0.11),
Gemma-4-E2B (0.17), Gemma-3-1B (0.24), and Gemma-3-4B (0.37).

\subsection{ActSVD}
\label{sec:results-actsvd}

Mean subspace overlap (rank $r=10$, mean squared cosine of principal
angles) is 0.28 for Gemma-3-1B, 0.26 for Gemma-3-4B, 0.20 for
Gemma-4-E2B, 0.15 for Llama-3.2-3B, and 0.18 for Qwen-3.5-4B
\ref{sec:figures-actsvdcsvd}. The overlap score is non-negative by
construction, since it averages squared cosines, so unlike DoM it
does not distinguish alignment from anti-alignment --- only their
combined magnitude. Three of the five models fall under 0.20, which
we read as ``largely orthogonal''; the two Gemma-3 checkpoints land
in the moderate-overlap range $0.20{-}0.50$.

Depth structure is consistent across all five models: overlap is
highest in the early third, drops through the middle, and reaches
its minimum in the late third. Llama-3.2-3B moves from 0.28 (early)
to 0.13 (middle) to 0.07 (late); Qwen-3.5-4B from 0.34 to 0.10 to
0.09; Gemma-4-E2B from 0.33 to 0.13 to 0.12; the two Gemma-3
checkpoints follow the same shape from $\sim$0.45 down to
$\sim$0.13--0.17. The most-separated layer in every model is in the
final third --- specifically layer 26 of 27, 34 of 35, 34 of 36,
28 of 29, and 31 of 33 --- with overlap scores of 0.13, 0.09, 0.05,
0.04, and 0.07 respectively. By the end of the network, harmful
information and helpful information occupy nearly perpendicular
subspaces of the residual stream.

The top principal-angle cosine tells the same story. Minimum top
cosines across post-embedding layers are 0.59, 0.65, 0.51, 0.49,
and 0.55 for the five models. No late-layer subspace pair exhibits
a single shared direction approaching cosine 1; the orthogonality
is distributed across the rank-10 subspace rather than driven by a
small number of dominant axes.

Ranking the five models from most to least orthogonal under
ActSVD (by mean overlap): Llama-3.2-3B (0.15), Qwen-3.5-4B (0.18),
Gemma-4-E2B (0.20), Gemma-3-4B (0.26), and Gemma-3-1B (0.28).

\subsection{DoM and ActSVD: Agreement}
\label{sec:results-synthesis}
The two methods agree qualitatively. DoM finds safety and utility
directions whose mean cosine is small in either sign across layers,
often more orthogonal than chance, with the partial exception of a
single anti-aligned mid-layer in Gemma-3-4B. ActSVD finds subspaces
whose principal angles increase with depth, becoming nearly
perpendicular in late layers. The model rankings are also nearly
identical between methods: Llama-3.2-3B and Qwen-3.5-4B show the
cleanest separation under both methods, while the two Gemma-3
checkpoints show the most residual entanglement. Both methods
identify the late layers as the cleanest candidates for
projection-style intervention, motivating the ablation experiments
in Section~\ref{sec:Causal Ablations}.


\subsection{Causal Ablations}
\label{sec:results-ablation}

Table~\ref{tab:ablation} reports AdvBench refusal rates and HelpSteer
helpfulness scores for all five conditions on Llama-3.2-3B, chosen
because it exhibits the cleanest geometric separation under both DoM
and ActSVD.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcc}
\hline
\textbf{Condition} & \textbf{AdvBench} & \textbf{HelpSteer} \\
                   & \textbf{Refusal (\%)} & \textbf{(1--5)} \\
\hline
Baseline       & 94.62 & 3.20 \\
Ablate safety  & 72.12 & 3.16 \\
Ablate utility & 98.08 & 2.63 \\
Ablate both    & 75.85 & 2.55 \\
Ablate random  & 96.92 & 3.25 \\
\hline
\end{tabular}
\caption{Causal ablation results on Llama-3.2-3B. AdvBench measures
safety (higher = more refusals); HelpSteer measures helpfulness
(higher = more helpful).}
\label{tab:ablation}
\end{table}

Ablating the safety direction reduces the refusal rate from 94.62\%
to 72.12\%, a 22.5 percentage-point drop, while HelpSteer score
remains nearly unchanged (3.16 vs.\ 3.20). This dissociation is
the central causal finding: removing the safety direction costs the
model roughly a quarter of its refusal behavior but does not
measurably impair helpfulness.

Ablating the utility direction produces the complementary pattern.
The refusal rate rises slightly to 98.08\% while HelpSteer drops substantially to 2.63,
a 0.57-point decline relative to baseline. Helpfulness and refusal
thus respond to opposite interventions, which is exactly what causal
separability predicts.

The ablate-both condition combines both hooks simultaneously. Its
refusal rate of 75.85\% closely tracks the safety-ablation result
(72.12\%), and its HelpSteer of 2.55 closely tracks the
utility-ablation result (2.63), consistent with the two effects
being approximately additive rather than interacting.

The control condition (ablate-random) is decisive. Projecting out a
random unit vector at the same layer leaves the refusal rate at
96.92\% and HelpSteer at 3.25, both within noise of baseline. The
behavioral changes produced by the safety and utility ablations are
therefore specific to the content of those directions and cannot be
attributed to the act of projection itself.

Together, the four conditions confirm that the geometric orthogonality
observed in Sections~\ref{sec:results-dom} and~\ref{sec:results-actsvd}
has causal force: safety and utility are encoded along directions that
can be surgically targeted without collateral damage to the other
behavior.

\label{sec:Causal Ablations}
